Optimal Finite Element Integration

### 2015-08-06 TOMS_Paper/assembly.tex
Several minor fixes

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or "flops", than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation systems or synthesizers. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning the on-going effort to produce a novel form compiler in FEniCS, called UFLACS (\cite{Uflacs}), which adds to the abundant set of code generation (or transformation) systems for assembly operators.

@@

We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are, respectively, a bilinear and a linear form. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are "assembled" and subsequently used to solve the linear system through (typically) an iterative method.

We focus on the assembly phase, which is often characterized as a two-step procedure: \textit{local} and \textit{global} assembly. Local assembly is the subject of this article. It consists of computing the contributions of a single element in the discretized domain to the equation solution. In global assembly, such local contributions are "coupled" by suitably inserting them into $A$ and $b$.

@@

\subsection{Tensor Contraction Mode}
Starting from Equation~\ref{eq:quadrature}, exploiting linearity, associativity and distributivity of the involved mathematical operators, we can rewrite the expression as
\begin{equation}
\label{eq:tensor}
A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
\end{equation}
A generalization of this transformation has been proposed in~\cite{Kirby:TC}. By only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables. The evaluation of the local tensor can then be abstracted as
\begin{equation}
A_{ij}^K = \sum_{\alpha} A_{i_1 i_2 \alpha}^0 G_{K}^\alpha
\end{equation}
in which the pre-evaluated "reference tensor" $A_{i_1 i_2 \alpha}$ and the cell-dependent "geometry tensor" $G_{K}^\alpha$ are exposed.

\subsection{Qualitative Comparison}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for "complex" forms. On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains "sufficiently small". The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-Compiler}). In this compiler, a heuristic is used to choose the most suitable mode for a given form. It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically (\cite{Fenics}). We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}. For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of reasoning at the monomials level -- not at the form level -- without affecting the code generation cost.

@@

A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of "some level of local memory" (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). We will provide precise details about this in the following sections. For the moment, and just as an example, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporary arrays of size $N\cdot O$; with some discretizations, this can break the local memory threshold.

@@

To answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with the input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated into temporaries. As part of this pre-evaluation, the reduction induced by $L_i$ vanishes. In the following, we simply refer to this special sort of code hoisting as "\textit{pre-evaluation}".

@@

The point of departure consists of understanding the impact, as number of operations saved or introduced, of pre-evaluation. This is studied "locally"; that is, for each monomial, in isolation. If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed. The result is a set of $n$-dimensional tables (these can be seen as "slices" of the reference tensor at the math level), $n$ being the arity of the multilinear form. Identical tables are mapped to the same temporary. Eventually, sharing is removed from the resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}. The transformed loop nest is as in Figure~\ref{code:loopnest-opt}.

@@

Before elaborating on the profitability of pre-evaluation, we need to discuss under which conditions this approach, based on a "local analysis" of monomials, is optimal.

\begin{Prop}
\label{prop:optimal-approach}
Consider an expression comprising a set of monomials $M$. Let $P$ be the set of pre-evaluated monomials, determined as described in Figure~\ref{code:intuition}, and be $Z = M \setminus P$. Assume that:
\begin{enumerate}
\item the cost function employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
\item pre-evaluating distinct monomials does not produce identical tables
\item monomials in $P$ do not share terms
\end{enumerate} 
Then, the loop nest $LN$ is optimal under memory constraints $C$, once sharing is removed.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) The creation of an optimal cost function is discussed in Section~\ref{sec:op_count}. 2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions. This is however a pathological case. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. For simplicity, we do not handle this case, which otherwise would require a "global" analysis of the monomials in the form. Further, even though, in practice, assumptions 2) and 3) did not hold, the resulting loop nest would still be nearly-optimal.

@@

B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, being the cost function optimal due to assumption 1). Assumption 3) guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would result sub-optimal due to destroying sharing-removal opportunities. The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$. This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are "de-classed" from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so adding $p$ to $Z$ does not augment the operation count. 
\end{proof}

@@

\subsection{Automation through the COFFEE Language}
As opposed to what happens in FFC with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general: all optimizations are expressed as composition of parametric "building-block" operations. This has several advantages. Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could ideally be expressed using the existing operators, or with small effort building on what is already available. Secondly, generality: other domains sharing properties similar to that of finite element integration (e.g., multilinear loop nests) could be optimized through the same compiler. Thirdly, robustness: the same building-block operations are exploited, and therefore stressed, by different optimization pipelines.

@@

COFFEE is agnostic with respect to the high level form compiler, so the first step consists of removing redundant sub-expressions. This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to trivial code without performing any sort of transformation. Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent to $L_i$. We then synthesize an optimal loop nest as described in the previous sections. This may in turn expose a set of $L_i$-dependent statements evaluating some temporaries. For each of this statements, we try to remove sharing by greedily applying factorization and code motion. In the COFFEE language, this process is expressed by simply concatenating five operators (i.e., five function calls named "factorize", "expand", etc.).

@@

% say that the reasons to have zeros is to keep the design of the form compiler simple, which just "puts zeros" and "outline the assembly expression".
% rationale: contiguous in the innermost dimension are aggregated. contiguous in all other dimensions are not, unless complete overlap.
% tables shrunk and merged

@@

We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix ("\texttt{Mass}") and the bilinear forms arising in a Helmholtz equation ("\texttt{Helmholtz}"), in an elastic model ("\texttt{Elasticity}"), and in a hyperelastic model ("\texttt{Hyperelasticity}"). The complete specification of these forms is made publicly available (\cite{ufl-code}).

We evaluate the speed ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
\begin{itemize}
\item the FEniCS Form Compiler's built-in optimization systems, namely
\begin{itemize}
\item optimized quadrature mode, \texttt{quad} (work presented in~\citeN{quadrature1})
\item tensor mode, \texttt{tens} (work presented in~\citeN{FFC-TC})
\item automatic mode, \texttt{auto} (here, heuristics are used to infer the optimal mode for the form between quadrature and tensor)
\end{itemize}
\item a novel back-end for the FEniCS Form Compiler, UFLACS, referred to as \texttt{ufls}, whose primary goals are improved code generation time and runtime
\item the sole generalized loop-invariant code motion in COFFEE plus padding, \texttt{cfO1}, as discussed in~\citeN{Luporini}
\item the optimal multilinear loop nest synthesis plus padding and symbolic execution, \texttt{cfO2}, which is the focus of this paper
\end{itemize}

The values that we report are the average of three runs with "warm cache" (no code generation time, no compilation time). They include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh used to run the simulations was chosen small enough to fit the L3 cache of the CPU, so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.

@@

Following the methodology (and the notation) used in ~\citeN{quadrature1} and~\citeN{Francis}, we increasingly vary the complexity of each form. In particular:
\begin{itemize}
\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test and trial functions always have same degree
\item the polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
Constants of our analysis are instead
\begin{itemize}
\item the space of test, trial, and coefficient functions, Lagrange
\item the mesh, tetrahedral with a total of 4374 elements
\end{itemize}

The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are "nested" plots. We refer to the outer plot as the "grid", while the inner are the actual "plots". In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis. The top-left plot in a grid shows the speed up over unmodified code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on. The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$. A grid can therefore be looked at from different perspectives, which allows us to make structured considerations on the performance achieved. In each plot there are six groups of bars, each group referring to a particular code variant (\texttt{quad, tens, ...}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

@@

A second observation concerns the "triggering" of pre-evaluation in \texttt{cfO2}. The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed. This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.

@@

\paragraph{Elasticity}
\begin{figure}
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model.}\label{fig:elasticity}
\end{figure}
...

>>>>

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation systems or synthesizers. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning the on-going effort to produce a novel form compiler in FEniCS, called UFLACS (\cite{Uflacs}), which adds to the abundant set of code generation (or transformation) systems for assembly operators.

@@

We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are, respectively, a bilinear and a linear form. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are ``assembled'' and subsequently used to solve the linear system through (typically) an iterative method.

We focus on the assembly phase, which is often characterized as a two-step procedure: \textit{local} and \textit{global} assembly. Local assembly is the subject of this article. It consists of computing the contributions of a single element in the discretized domain to the equation solution. In global assembly, such local contributions are ``coupled'' by suitably inserting them into $A$ and $b$.

@@

\subsection{Tensor Contraction Mode}
Starting from Equation~\ref{eq:quadrature}, exploiting linearity, associativity and distributivity of the involved mathematical operators, we can rewrite the expression as
\begin{equation}
\label{eq:tensor}
A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
\end{equation}
A generalization of this transformation has been proposed in~\cite{Kirby:TC}. By only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables. The evaluation of the local tensor can then be abstracted as
\begin{equation}
A_{ij}^K = \sum_{\alpha} A_{i_1 i_2 \alpha}^0 G_{K}^\alpha
\end{equation}
in which the pre-evaluated ``reference tensor'' $A_{i_1 i_2 \alpha}$ and the cell-dependent ``geometry tensor'' $G_{K}^\alpha$ are exposed.

\subsection{Qualitative Comparison}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for ``complex'' forms. On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''. The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-TC}). In this compiler, a heuristic is used to choose the most suitable mode for a given form. It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically (\cite{Fenics}). We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}. For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of reasoning at the monomials level -- not at the form level -- without affecting the code generation cost.

@@

A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). We will provide precise details about this in the following sections. For the moment, and just as an example, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporary arrays of size $N\cdot O$; with some discretizations, this can break the local memory threshold.

@@

To answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with the input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated into temporaries. As part of this pre-evaluation, the reduction induced by $L_i$ vanishes. In the following, we simply refer to this special sort of code hoisting as ``\textit{pre-evaluation}''.

@@

The point of departure consists of understanding the impact, as number of operations saved or introduced, of pre-evaluation. This is studied ``locally''; that is, for each monomial, in isolation. If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form. Identical tables are mapped to the same temporary. Eventually, sharing is removed from the resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}. The transformed loop nest is as in Figure~\ref{code:loopnest-opt}.

@@

Before elaborating on the profitability of pre-evaluation, we need to discuss under which conditions this approach, based on a ``local analysis'' of monomials, is optimal.

\begin{Prop}
\label{prop:optimal-approach}
Consider an expression comprising a set of monomials $M$. Let $P$ be the set of pre-evaluated monomials, determined as described in Figure~\ref{code:intuition}, and be $Z = M \setminus P$. Assume that:
\begin{enumerate}
\item the cost function employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
\item pre-evaluating distinct monomials does not produce identical tables
\item monomials in $P$ do not share terms
\end{enumerate} 
Then, the loop nest $LN$ is optimal under memory constraints $C$, once sharing is removed.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) The creation of an optimal cost function is discussed in Section~\ref{sec:op_count}. 2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions. This is however a pathological case. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. For simplicity, we do not handle this case, which otherwise would require a ``global'' analysis of the monomials in the form. Further, even though, in practice, assumptions 2) and 3) did not hold, the resulting loop nest would still be nearly-optimal.

@@

B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, being the cost function optimal due to assumption 1). Assumption 3) guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would result sub-optimal due to destroying sharing-removal opportunities. The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$. This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so adding $p$ to $Z$ does not augment the operation count. 
\end{proof}

@@

\subsection{Automation through the COFFEE Language}
As opposed to what happens in FFC with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general: all optimizations are expressed as composition of parametric ``building-block'' operations. This has several advantages. Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could ideally be expressed using the existing operators, or with small effort building on what is already available. Secondly, generality: other domains sharing properties similar to that of finite element integration (e.g., multilinear loop nests) could be optimized through the same compiler. Thirdly, robustness: the same building-block operations are exploited, and therefore stressed, by different optimization pipelines.

@@

COFFEE is agnostic with respect to the high level form compiler, so the first step consists of removing redundant sub-expressions. This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to trivial code without performing any sort of transformation. Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent to $L_i$. We then synthesize an optimal loop nest as described in the previous sections. This may in turn expose a set of $L_i$-dependent statements evaluating some temporaries. For each of this statements, we try to remove sharing by greedily applying factorization and code motion. In the COFFEE language, this process is expressed by simply concatenating five operators (i.e., five function calls named ``factorize'', ``expand'', etc.).

@@

% say that the reasons to have zeros is to keep the design of the form compiler simple, which just ``puts zeros'' and ``outline the assembly expression''.
% rationale: contiguous in the innermost dimension are aggregated. contiguous in all other dimensions are not, unless complete overlap.
% tables shrunk and merged

@@

We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}''). The complete specification of these forms is made publicly available (\cite{ufl-code}).

We evaluate the speed ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
\begin{itemize}
\item FEniCS Form Compiler's built-in optimization system: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler's built-in optimization system: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler's built-in optimization system: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, discussed in~\citeN{Fenics}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion and padding (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal multilinear loop nest synthesis, padding and symbolic execution (captures the contributions of this paper). Referred to as \texttt{cfO2} 
\end{itemize}

The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time). They include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh used to run the simulations was chosen small enough to fit the L3 cache of the CPU, so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.

@@

Since we are running sequential programs, with respect to the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, we set $T_H = \infty$; that is, temporaries can occupy an arbitrary amount of memory. This is because a single process that has exclusive access to all of the bandwidth available in the system can count on pre-fetching to minimize cache misses. If, on the other hand, a parallel program were run, then a more sensible value for $T_H$ would be the size of the last level of private cache; this would avoid memory bank conflicts and, therefore, potential drops in runtime.

Following the methodology adopted in ~\citeN{quadrature1}, we increasingly vary the complexity of each form. In particular:
\begin{itemize}
\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test and trial functions always have same degree
\item the polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
Constants of our analysis are instead
\begin{itemize}
\item the space of test, trial, and coefficient functions, Lagrange
\item the mesh, tetrahedral with a total of 4374 elements
\end{itemize}

The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are ``nested'' plots. We refer to the outer plot as the ``grid'', while the inner are the actual ``plots''. In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis. The top-left plot in a grid shows the speed up over unmodified code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on. The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$. A grid can therefore be looked at from different perspectives, which allows us to make structured considerations on the performance achieved. In each plot there are six groups of bars, each group referring to a particular code variant (\texttt{quad, tens, ...}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

@@

A second observation concerns the ``triggering'' of pre-evaluation in \texttt{cfO2}. The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed. This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.

@@

\paragraph{Elasticity}
\begin{figure}
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model.}\label{fig:elasticity}
\end{figure}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. We note how the fact that\texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present.
